### 2024-10-10 manuscript.tex
conclusions/future perspectives: fix [ci skip]
This build is based on
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/66a49f4f63bfcb89ba3bd7ab5a0a44478728a6ec.

This commit was created by the following CI build and job:
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/66a49f4f63bfcb89ba3bd7ab5a0a44478728a6ec/checks
https://github.com/pivlab/annual_review_of_biomedical_data_science/actions/runs/11280057046

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/c2524cc88cbe2714df69644fdc4b9d3d5105b835/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/c2524cc88cbe2714df69644fdc4b9d3d5105b835}{pivlab/annual\_review\_of\_biomedical\_data\_science@c2524cc}
on October 10, 2024.

@@

\hypertarget{future-perspectives}{%
\subsection{Future perspectives}\label{future-perspectives}}

\hypertarget{conclusions}{%
\subsection{Conclusions}\label{conclusions}}

@@

>>>>

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/66a49f4f63bfcb89ba3bd7ab5a0a44478728a6ec/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/66a49f4f63bfcb89ba3bd7ab5a0a44478728a6ec}{pivlab/annual\_review\_of\_biomedical\_data\_science@66a49f4}
on October 10, 2024.

@@

\hypertarget{conclusions-and-future-perspectives}{%
\subsection{Conclusions and future perspectives}\label{conclusions-and-future-perspectives}}

Genomics plays an important role in realizing the promise of personalized medicine, although not without significant challenges.
In this review, we discuss different molecular phenotypes, theoretical models and computational approaches that offer different mechanistic views on how genetic variation leads to phenotypic changes.
To fully understand the molecular basis of human complex traits, it is essential to integrate diverse datasets and unravel the complex structures of biological networks.
Gene co-expression networks, represented as gene modules, have been successfully used across a variety of fields {[}\protect\hyperlink{ref-CqvNaEqc}{92},\protect\hyperlink{ref-r8alvfcC}{93},\protect\hyperlink{ref-E1tpJnqG}{94}{]}.
Here, we reviewed the main components of a gene module-based approach that leverages gene co-expression patterns learned from large transcriptomic datasets to integrate GWAS, TWAS, and drug-induced transcriptional profiles.
Drawing on modern theories of the genetic architecture of complex traits, we show that these approaches can prioritize core genes and create interpretable frameworks for drug repurposing.

Gene-gene interactions are a key component in the omnigenic model.
While the integration of gene modules with gene-trait associations and gene-drug data has been a valuable first step, it represents only a fraction of the complexity inherent in biological systems.
Biological networks extend far beyond these initial layers, involving intricate, multilayered systems such as histone modifications, transcription factor binding sites, \emph{cis}-regulatory networks, and chromatin conformation---all of which contribute to the co-regulation of gene transcription.
Moreover, this complexity includes protein-protein interaction networks, which form the backbone of signaling pathways and facilitate the transmission of signals both within and between cells.
Additionally, metabolic networks influence enzyme activities and metabolite levels, propagating changes that affect numerous other molecules {[}\protect\hyperlink{ref-dg9nKuy0}{50},\protect\hyperlink{ref-KNAOTAKP}{95}{]}.
Incorporating and quantifying these additional biological layers will enhance gene module construction, offering a more comprehensive and accurate representation of the intricate nature of real biological networks.

A gene module approach that integrates genetic studies has several important inputs: 1) gene expression data relevant to the research problem, such as large heterogeneous datasets that span over different conditions, or otherwise focused, smaller dataset that capture more subtle, unique, and disease-specific gene modules; 2) unbiased and relevant prior information that can be used during or after module extraction to distinguish relevant transcriptomic signatures from technical noise; and 3) reliable single variant/gene associations with the complex traits of interest.
Therefore, it is important to emphasize that a gene module approach is by no means a replacement of current single variant/gene approaches.
The development of new and improved statistical approaches that detect more causal genes, regardless of whether they have a direct or indirect effect on the trait, while reducing false positives {[}\protect\hyperlink{ref-i8Qjy4pw}{96}{]} will only improve other approaches that rely on them, such as gene module-based methods described here.
In this review, we show that the identification of peripheral genes is key to prioritize core genes that might represent more attractive drug targets, strongly suggesting that the mechanism proposed by the omnigenic model, i.e., that gene regulatory networks are highly interconnected and genes have specific roles with more direct or indirect effects on traits, is useful.

As the demand for analyzing large datasets and uncovering complex biological patterns grows, the need for faster and more efficient computational methods becomes increasingly critical.
Future bioinformatics tools must not only handle the scale of these tasks but also ensure that they are accessible and adaptable across different computational environments.
To meet these challenges, we are focusing on accelerating workflows with graphics processing unit (GPU) technology and exploring new platforms, such as WebAssembly, to bring powerful tools directly to users' browsers.

Leveraging GPU acceleration holds great promise for significantly speeding up bioinformatics workflows.
Research has shown that GPU technology can provide remarkable performance boosts, with speed improvements reaching up to 1000x for specific tasks {[}\protect\hyperlink{ref-1EIbaYHGq}{97}{]}.
Integrating GPU-aware libraries like CuPy {[}\protect\hyperlink{ref-JMQXR7X9}{98}{]} and RAPIDS {[}\protect\hyperlink{ref-wPsNXL1Q}{99}/{]} into our tools allows for seamless performance upgrades in established frameworks such as NumPy {[}\protect\hyperlink{ref-1LexfAxj}{100}{]} and Scikit-learn {[}\protect\hyperlink{ref-AujvwLp6}{101}{]}.
As our tools evolve, we anticipate developing custom GPU kernels, further enhancing speed and optimizing them for specific computational needs.
Advanced GPU programming with PyCUDA {[}\protect\hyperlink{ref-b5xmRoLZ}{102}{]} could also be key to future optimizations.

Another promising direction is utilizing WebAssembly {[}\protect\hyperlink{ref-ITQaZvrQ}{103}{]} to make bioinformatics tools more portable and accessible.
We aim to adapt existing bioinformatics applications into web-based tools that can be run directly in the browser, providing users with full control over their data and eliminating the need for data transfer over networks.
Projects like Biowasm {[}\protect\hyperlink{ref-Z0f406Zs}{104}/{]}, Kana {[}\protect\hyperlink{ref-1DswCJYT0}{105}{]}, and VirtualWasm {[}\protect\hyperlink{ref-WhJVE7a4}{106}{]} have already demonstrated the potential of WebAssembly for genomics and single-cell analysis, and we plan to build on this foundation.
By executing statistical analyses, computations, and visualizations directly on the user's device, WebAssembly will not only address privacy and compliance concerns but also reduce server costs and computational load by shifting the workload to the client side.

Future innovations in GPU acceleration and WebAssembly will transform how bioinformatics tools are developed and deployed, leading to faster, more accessible solutions for researchers worldwide.
By integrating these technologies, we can build the next generation of computational platforms that combine speed, flexibility, and privacy, unlocking new potential for biological discoveries.
As these technologies mature, they will enable even more ambitious and large-scale analyses to be performed quickly and efficiently.

@@

\leavevmode\vadjust pre{\hypertarget{ref-CqvNaEqc}{}}%
\CSLLeftMargin{92. }%
\CSLRightInline{\textbf{WGCNA: an R package for weighted correlation network analysis}
\CSLBlock{Peter Langfelder, Steve Horvath} \emph{BMC Bioinformatics} (2008-12) \url{https://doi.org/bwdbzc}
\CSLBlock{DOI: \href{https://doi.org/10.1186/1471-2105-9-559}{10.1186/1471-2105-9-559} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/19114008}{19114008} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2631488}{PMC2631488}}}

\leavevmode\vadjust pre{\hypertarget{ref-r8alvfcC}{}}%
\CSLLeftMargin{93. }%
\CSLRightInline{\textbf{WGCNA Application to Proteomic and Metabolomic Data Analysis}
\CSLBlock{G Pei, L Chen, W Zhang} \emph{Methods in Enzymology} (2017) \url{https://doi.org/grksc8}
\CSLBlock{DOI: \href{https://doi.org/10.1016/bs.mie.2016.09.016}{10.1016/bs.mie.2016.09.016} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/28109426}{28109426}}}

\leavevmode\vadjust pre{\hypertarget{ref-E1tpJnqG}{}}%
\CSLLeftMargin{94. }%
\CSLRightInline{\textbf{Identification of glioblastoma~gene prognosis modules based on weighted gene co-expression network analysis}
\CSLBlock{Pengfei Xu, Jian Yang, Junhui Liu, Xue Yang, Jianming Liao, Fanen Yuan, Yang Xu, Baohui Liu, Qianxue Chen} \emph{BMC Medical Genomics} (2018-11-01) \url{https://doi.org/gfzt4j}
\CSLBlock{DOI: \href{https://doi.org/10.1186/s12920-018-0407-1}{10.1186/s12920-018-0407-1} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/30382873}{30382873} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6211550}{PMC6211550}}}

\leavevmode\vadjust pre{\hypertarget{ref-KNAOTAKP}{}}%
\CSLLeftMargin{95. }%
\CSLRightInline{\textbf{Boosting GWAS using biological networks: A study on susceptibility to familial breast cancer}
\CSLBlock{Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott} \emph{PLOS Computational Biology} (2021-03-18) \url{https://doi.org/g7zdns}
\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pcbi.1008819}{10.1371/journal.pcbi.1008819} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/33735170}{33735170} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8009366}{PMC8009366}}}

\leavevmode\vadjust pre{\hypertarget{ref-i8Qjy4pw}{}}%
\CSLLeftMargin{96. }%
\CSLRightInline{\textbf{Adjusting for genetic confounders in transcriptome-wide association studies improves discovery of risk genes of complex traits}
\CSLBlock{Siming Zhao, Wesley Crouse, Sheng Qian, Kaixuan Luo, Matthew Stephens, Xin He} \emph{Nature Genetics} (2024-01-26) \url{https://doi.org/gtf7kb}
\CSLBlock{DOI: \href{https://doi.org/10.1038/s41588-023-01648-9}{10.1038/s41588-023-01648-9} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/38279041}{38279041} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10864181}{PMC10864181}}}

\leavevmode\vadjust pre{\hypertarget{ref-1EIbaYHGq}{}}%
\CSLLeftMargin{97. }%
\CSLRightInline{\textbf{Graphics processing units in bioinformatics, computational biology and systems biology}
\CSLBlock{Marco S Nobile, Paolo Cazzaniga, Andrea Tangherloni, Daniela Besozzi} \emph{Briefings in Bioinformatics} (2016-07-08) \url{https://doi.org/g7zg6w}
\CSLBlock{DOI: \href{https://doi.org/10.1093/bib/bbw058}{10.1093/bib/bbw058} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/27402792}{27402792} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5862309}{PMC5862309}}}

\leavevmode\vadjust pre{\hypertarget{ref-JMQXR7X9}{}}%
\CSLLeftMargin{98. }%
\CSLRightInline{\textbf{cupy/cupy}
\CSLBlock{CuPy} (2024-10-10) \url{https://github.com/cupy/cupy}}

\leavevmode\vadjust pre{\hypertarget{ref-wPsNXL1Q}{}}%
\CSLLeftMargin{99. }%
\CSLRightInline{\textbf{RAPIDS \textbar{} GPU Accelerated Data Science}
\CSLBlock{RAPIDS \textbar{} GPU Accelerated Data Science} \url{https://rapids.ai/}}

\leavevmode\vadjust pre{\hypertarget{ref-1LexfAxj}{}}%
\CSLLeftMargin{100. }%
\CSLRightInline{\textbf{Array programming with NumPy}
\CSLBlock{Charles R Harris, KJarrod Millman, Stéfan J van der Walt, Ralf Gommers, Pauli Virtanen, David Cournapeau, Eric Wieser, Julian Taylor, Sebastian Berg, Nathaniel J Smith, \ldots{} Travis E Oliphant} \emph{Nature} (2020-09-16) \url{https://doi.org/ghbzf2}
\CSLBlock{DOI: \href{https://doi.org/10.1038/s41586-020-2649-2}{10.1038/s41586-020-2649-2} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/32939066}{32939066} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7759461}{PMC7759461}}}

\leavevmode\vadjust pre{\hypertarget{ref-AujvwLp6}{}}%
\CSLLeftMargin{101. }%
\CSLRightInline{\textbf{Scikit-learn: Machine Learning in Python}
\CSLBlock{Fabian Pedregosa, Gaël Varoquaux, Alexandre Gramfort, Vincent Michel, Bertrand Thirion, Olivier Grisel, Mathieu Blondel, Andreas Müller, Joel Nothman, Gilles Louppe, \ldots{} Édouard Duchesnay} \emph{arXiv} (2012) \url{https://doi.org/gpptwn}
\CSLBlock{DOI: \href{https://doi.org/10.48550/arxiv.1201.0490}{10.48550/arxiv.1201.0490}}}

\leavevmode\vadjust pre{\hypertarget{ref-b5xmRoLZ}{}}%
\CSLLeftMargin{102. }%
\CSLRightInline{\textbf{PyCUDA and PyOpenCL: A scripting-based approach to GPU run-time code generation}
\CSLBlock{Andreas Klöckner, Nicolas Pinto, Yunsup Lee, Bryan Catanzaro, Paul Ivanov, Ahmed Fasih} \emph{Parallel Computing} (2012-03) \url{https://doi.org/c8hxzm}
\CSLBlock{DOI: \href{https://doi.org/10.1016/j.parco.2011.09.001}{10.1016/j.parco.2011.09.001}}}

\leavevmode\vadjust pre{\hypertarget{ref-ITQaZvrQ}{}}%
\CSLLeftMargin{103. }%
\CSLRightInline{\textbf{Bringing the web up to speed with WebAssembly}
\CSLBlock{Andreas Haas, Andreas Rossberg, Derek L Schuff, Ben L Titzer, Michael Holman, Dan Gohman, Luke Wagner, Alon Zakai, JF Bastien} \emph{Proceedings of the 38th ACM SIGPLAN Conference on Programming Language Design and Implementation} (2017-06-14) \url{https://doi.org/gfgvf4}
\CSLBlock{DOI: \href{https://doi.org/10.1145/3062341.3062363}{10.1145/3062341.3062363}}}

\leavevmode\vadjust pre{\hypertarget{ref-Z0f406Zs}{}}%
\CSLLeftMargin{104. }%
\CSLRightInline{\textbf{Biowasm} \url{https://biowasm.com/}}

\leavevmode\vadjust pre{\hypertarget{ref-1DswCJYT0}{}}%
\CSLLeftMargin{105. }%
\CSLRightInline{\textbf{Powering single-cell analyses in the browser with WebAssembly}
\CSLBlock{Aaron Lun, Jayaram Kancherla} \emph{Cold Spring Harbor Laboratory} (2022-03-04) \url{https://doi.org/gtk3dc}
\CSLBlock{DOI: \href{https://doi.org/10.1101/2022.03.02.482701}{10.1101/2022.03.02.482701}}}

\leavevmode\vadjust pre{\hypertarget{ref-WhJVE7a4}{}}%
\CSLLeftMargin{106. }%
\CSLRightInline{\textbf{ViralWasm: a client-side user-friendly web application suite for viral genomics}
\CSLBlock{Daniel Ji, Robert Aboukhalil, Niema Moshiri} \emph{Bioinformatics} (2024-01-01) \url{https://doi.org/g7zg6x}
\CSLBlock{DOI: \href{https://doi.org/10.1093/bioinformatics/btae018}{10.1093/bioinformatics/btae018} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/38200583}{38200583} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10809900}{PMC10809900}}}
